Forecasting 2024 US election using linear regression

### 2024-10-28 paper/paper.qmd
edit on model justification

---
title: "Forecasting 2024 US election using linear regression"
author: 
  - Diana Shen
  - Jinyan Wei
  - Jerry Yu
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Forecast-2024-US-election](https://github.com/DianaShen1224/Forecast-2024-US-election)."
date: today
date-format: long
abstract: "This paper predicts the outcome of the 2024 U.S. presidential election using a statistical model based on aggregated polling data for Kamala Harris and Donald Trump. We employ multi-level regression with post-stratification (MRP) using demographic predictors to estimate voter support. The analysis addresses polling biases and proposes an idealized survey methodology with a $100,000 budget. Our results offer insights into voter behavior and suggest improvements for future election forecasting models."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

We performed multi-level regression with post-stratification (MRP) to predict support for Kamala Harris in the upcoming election. To perform MRP, we fit a logistic regression model to predict support for Harris using sex, age, race, race_hispanic, highest level of education, state, and urban as predictors on our survey data set [@americaspoliticalpulse], and post-stratified it using ACS census data [@citeIPUMS].

## Model set-up

We built our Bayesian logistic regression model using the `stan_glm()` function of the `rstanarm` package [@rstanarm]. The model that we use is:

```{=tex}
\begin{align} 
vote\_harris_i|\pi_i &\sim \text{Bernoulli}(\pi_i) \\
\text{logit}(\pi_i) &= \beta_0 + \beta_1 \text{sex}_i + \beta_2  \text{age\_bracket}_i + \beta_3 \text{race}_i + \beta_4 \text{race\_hispanic}_i + \beta_5 \text{education\_level}_i + \beta_6 \text{urban}_i \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\beta_4 &\sim \text{Normal}(0, 2.5) \\
\beta_5 &\sim \text{Normal}(0, 2.5) \\
\beta_6 &\sim \text{Normal}(0, 2.5)
\end{align}
```
where binary variable $vote\_harris_i$ is equal to 1 if the respondent's preferred 2024 presidential candidate is Kamala Harris, or 0 if the respondent's preferred 2024 presidential candidate is Donald Trump. We run our model in R [@citeR] using `stan_glm`, with the default priors from `rstanarm` [@citerstanarm]. We then apply our model to our post-stratification data [@citeIPUMS] to predict the popular vote and electoral college results of the 2024 U.S. Presidential election.

Our model uses logistic regression, therefore one of its weaknesses is the fact that it predicts a binary outcome and does not consider the possibility that some American adults might vote for a third-party candidate or abstain from voting. Tradeoffs and concerns about our model are discussed in more detail in @sec-weaknesses-limitations.

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between certain demographic characteristics and support for Kamala Harris. Specifically:

-   **Sex**: We might expect differences in support between male and female respondents, with women potentially being more likely to support Kamala Harris due to certain policy stances that may appeal to this demographic.
-   **Age Bracket**: Age can be a significant factor, as younger voters might lean more progressive, potentially increasing support for Kamala Harris compared to Donald Trump.
-   **Race and Hispanic Ethnicity**: Historical voting patterns suggest that non-white racial groups, including Hispanic voters, may show stronger support for Democratic candidates, such as Harris. This is reflected in the coefficients $\beta_3$ for race and $\beta_4$ for Hispanic ethnicity.
-   **Education Level**: Higher education levels are often correlated with progressive political views, so we might expect a positive relationship between education level and support for Harris.
-   **Urban**: Respondents living in urban areas may be more likely to support Kamala Harris due to the urban-rural political divide observed in recent elections, where urban areas tend to lean more Democratic.

The coefficients, such as $\beta_1$ for sex and $\beta_2$ for age bracket, represent the strength of these relationships. For example, a positive value of $\beta_1$ would suggest that being female increases the probability of supporting Harris, all else being equal.

By using normal priors like $\beta_1 \sim \text{Normal}(0, 2.5)$, we assume that, before seeing the data, the effect of each predictor on the log-odds of supporting Kamala Harris is centered around zero, but with flexibility for the data to inform the posterior estimates.

The Bayesian framework allows us to capture the uncertainty in our estimates and to refine our understanding of how different factors influence political support as new data becomes available. This is especially important in political forecasting, where the relationships between demographic factors and candidate support can be influenced by evolving social and political dynamics.

@@

The data cleaning process involved filtering data to just high-quality pollsters (i.e numeric_grade >= 3.0) and filter to just Harris (i.e. candidate_name == "Kamala Harris")

@@

| **Region**    | **States**                                                                                                                                                 |
|------------------------|------------------------------------------------|
| **MIDWEST**   | Illinois, Indiana, Iowa, Kansas, Michigan, Minnesota, Missouri, Nebraska, North Dakota, Ohio, South Dakota, Wisconsin                                      |
| **Northeast** | Connecticut, Delaware, District of Columbia, Maine, Maryland, Massachusetts, New Hampshire, New Jersey, New York, Pennsylvania, Rhode Island, Vermont      |
| **SOUTH**     | Alabama, Arkansas, Florida, Georgia, Kentucky, Louisiana, Mississippi, North Carolina, Oklahoma, South Carolina, Tennessee, Texas, Virginia, West Virginia |
| **WEST**      | Alaska, Arizona, California, Colorado, Hawaii, Idaho, Montana, Nevada, New Mexico, Oregon, Utah, Washington, Wyoming                                       |

>>>>

---
title: "Forecasting 2024 US election using linear regression"
author: 
  - Diana Shen
  - Jinyan Wei
  - Jerry Yu
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Forecast-2024-US-election](https://github.com/DianaShen1224/Forecast-2024-US-election)."
date: today
date-format: long
abstract: "This paper predicts the outcome of the 2024 U.S. presidential election using a statistical model based on aggregated polling data for Kamala Harris and Donald Trump. We employ multi-level regression with post-stratification (MRP) using demographic predictors to estimate voter support. The analysis addresses polling biases and proposes an idealized survey methodology with a $100,000 budget. Our results offer insights into voter behavior and suggest improvements for future election forecasting models."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

We conducted a regression analysis to predict support for Kamala Harris
in the 2024 U.S. presidential election. Our analysis includes two
models: an **unweighted linear model** and a **weighted linear model**.
These models predict Harris’s support as a continuous outcome across
different states, using aggregated polling data with a focus on
adjusting for poll quality, sample size, recency, and potential
partisanship biases. This setup aligns with the New York Times
methodology for election polling, which assigns weights based on poll
quality, sample size, and recency of the poll to reflect an accurate
picture of voter sentiment [@nytimes].

## Model Set-up

We implemented our linear regression models using the `lm()` function in
R. The unweighted model provides a baseline with equal treatment for all
polls, while the weighted model adjusts each poll observation based on
combined factors, including recency, sample size, and poll quality. The
model specifications are as follows:

#### Unweighted Linear Model

The unweighted model serves as a baseline, treating all polls equally
without incorporating adjustments for recency, sample size, or poll
quality. The model is defined as:

$$
\text{Support_Harris}_i = \beta_0 + \beta_1 \cdot \text{National_Poll}_i + \epsilon_i
$$

This unweighted model provides a baseline estimate of Harris’s support
across different polls without considering variations in poll recency,
sample size, or quality. It allows us to understand Harris’s support
level with a simplified assumption that all polls are of equal
relevance.

#### Weighted Linear Model

To enhance the accuracy of our predictions, the weighted model
incorporates adjustments for recency, sample size, and pollster quality.
This model is defined as:

$$
\text{Support_Harris}_i = \beta_0 + \beta_1 \cdot \text{National_Poll}_i + \beta_2 \cdot \text{Recency_Weight}_i + \beta_3 \cdot \text{Sample_Size_Weight}_i + \beta_4 \cdot \text{Numeric_Grade}_i + \epsilon_i
$$

The weighted model applies these additional factors to account for
variations in data quality and recency, offering a refined view of
Harris’s support.

## Model Justification

In developing our model to predict support for Kamala Harris, we
included adjustments based on national polling, recency, sample size,
and pollster grade. These features were selected to mitigate common
biases in polling data and to ensure our model reflects up-to-date and
high-quality information. The decision to incorporate recency as a
predictor is motivated by the observation that voter sentiment can shift
rapidly in response to events, such as debates or political
controversies, especially as the election date approaches (The New York
Times, 2024). Sample size was also used as a weight, as larger polls
tend to reduce sampling variability up to a reasonable threshold, set at
2,300 responses.

Pollster quality, represented through numeric grades, accounts for
systematic differences in methodology, professionalism, and past
accuracy. Pollsters with a strong historical performance are given more
weight, as they are more likely to produce unbiased and reliable
results. The model structure, however, is based on linear regression,
meaning it assumes an additive relationship between these predictors and
voter support. This simplification, while computationally efficient and
interpretable, may not capture more complex interactions between
factors, such as shifts in demographic or regional support.

Our model was implemented in R using the `lm()` function for the
unweighted model and a weighted version for the adjusted model.
Validation was performed using out-of-sample testing and RMSE
calculations to assess accuracy. Sensitivity analyses were also
conducted, examining how the model performed across different subsets of
data.

The model has several limitations, including the linearity assumption,
which may not fully capture the dynamic nature of voter support. Its
effectiveness relies on the assumption that the weights (recency, sample
size, pollster grade) adequately mitigate biases in polling data.
Additionally, it does not account for third-party candidates or
undecided voters, which could affect the election outcome. While
Bayesian hierarchical models were considered for their capability to
handle multi-level data, they were ultimately set aside for a simpler,
more interpretable linear model due to the analysis's scope and
computational constraints. Future research should explore these more
complex methods if time and resources permit.

@@

The data cleaning process involved filtering data to just high-quality
pollsters (i.e numeric_grade \>= 3.0) and filter to just Harris (i.e.
candidate_name == "Kamala Harris")

@@

| **Region**    | **States**                                                                                                                                                 |
|-------------------------|-----------------------------------------------|
| **MIDWEST**   | Illinois, Indiana, Iowa, Kansas, Michigan, Minnesota, Missouri, Nebraska, North Dakota, Ohio, South Dakota, Wisconsin                                      |
| **Northeast** | Connecticut, Delaware, District of Columbia, Maine, Maryland, Massachusetts, New Hampshire, New Jersey, New York, Pennsylvania, Rhode Island, Vermont      |
| **SOUTH**     | Alabama, Arkansas, Florida, Georgia, Kentucky, Louisiana, Mississippi, North Carolina, Oklahoma, South Carolina, Tennessee, Texas, Virginia, West Virginia |
| **WEST**      | Alaska, Arizona, California, Colorado, Hawaii, Idaho, Montana, Nevada, New Mexico, Oregon, Utah, Washington, Wyoming                                       |
